The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-07-02 paper/ms.tex
new version with updated discussion, concluiosns and notation in the analytical solution

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$ 
emission line in distant galaxies. 
We model a galaxy as a sphere with an homogeneous mixture of dust and
hydrogen at a constant temperature. 
These spheres have a solid-body rotation with maximum velocities in
the range $0-300$ \kms and neutral hydrogen optical depths in the
range $\tau_{\rm H}=10^{5}-10^{7}$. 
We consider two types of source distributions: single central and a
homogeneoulsy distributed in the sphere. 
Our main result is that rotation introduces a dependence of the
line morphology with  viewing angle. 
Observations with a line of sight parallel to the rotation axis yield
line morphologies similar to the static case. 
For lines of sight perpendicular to the rotation axis the intensity at the line
center and the line width increase with rotational velocity.   
Along the same sight, the line becomes single peaked  at rotational
velocities close to half the line width in the static case. 
Notably, we find that rotation does not induce any significant spatial
anisotropy in the integrated line flux, in the escape fraction or the
average number of scatterings. 
We also provide an analytical expression for the outgoing \lya spectrum of
a rotating cloud which accounts for the velocity and viewing angle
trends observed in our Monte Carlo simulations.
\end{abstract}

@@

For each photon in the simulation we have its initial position inside
the sphere, direction of propagation $\hat{k}_{\rm in}$ and reduced
frequency $x_{\rm in}$. 
The photon's propagation stops once they cross the
surface of the sphere. At this point we store the position, the outgoing direction
of propagation $\hat{k}_{\rm out}$ and the reduced frequency $x_{\rm
  out}$. We now define the angle $\theta$ by $\cos\theta = \hat{k}_{\rm out}\cdot
\hat{k}\equiv \mu$, it is the angle of the outgoing photons with 
respect to the direction of the angular velocity. We use the variable $\mu$ to
study the anisotropy induced by rotation. Figure \ref{fig:geometry}
shows the geometry of the problem and the important variables.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f3.pdf}
\end{center}
\caption{Same as Figure \ref{fig:CentralSpec} for \ly photons
  initialized homogeneously throughout the sphere.
    \label{fig:HomSpec}}  
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f4.pdf}
\end{center}
\caption{Shape of the \ly line for different maximum rotational
  velocities for a LoS perpendicular to the rotation axis
  ($|\mu|\sim 0$). The continuous (dashed) line represents the central
  (homogeneous) source distributions.  The panels follow the same
  distribution as in Figures \ref{fig:CentralSpec} and \ref{fig:HomSpec}.
    \label{fig:differentvelocities}}  
\end{figure*}

@@

The central result of this paper are summarized in Figures
\ref{fig:CentralSpec} and \ref{fig:HomSpec}. 
They show 2D histograms of the escape frequency $x$ and outgoing angle
$\theta$ parameterized by $|\mu|$. 
Taking into account only photons photons around a value
of $|\mu|$ gives us the emission detected by an observer located at an
angle $\theta$ with respect to the rotation axis. 
We have veryfied that the solutions are indeed symmetric with respect
to $\mu=0$.

From these Figures we can see that the line properties change with
rotational velocity and depend on the viewing angle $\theta$.  
In the next subsections we quantify the morphology changes with with
velocity, optical depth and viewing angle.  
We characterize the line morphology by its total intensity, the full
width at half maximu, (FWHM) and the location of the peak maxima. 
In order to interpret the
morphological changes in the line we also report the median number of
scatter for each \ly photon in the simulation. 
For the models where dust is included we measure the escape fraction
as a function of rotational velocity.

@@

The first column in both Figures \ref{fig:CentralSpec} and
\ref{fig:HomSpec} shows that for the static sphere the line properties
are independent of $|\mu|$, as it is expected due
to the spherical symmetry. 
However, for increasing rotational velocities, at a fixed optical
depth, there are clear signs that this symmetry is broken.

@@

This is clear in  Figure \ref{fig:differentvelocities} where we
present the different line morphologies for $|\mu|\sim 0$ in the
homogeneous and central configurations.  
The panels have the same distribution as Figures \ref{fig:CentralSpec}
and \ref{fig:HomSpec}.    
There are three clear effects on the line morphology as the rotational
velocity increases.  
First, the line broadens; second, the double peaks reduce their intensity; and
third, the intensity at the line centre rises. 
The last two effects are combined to give the impression that the double
peaks are merged into a single one at high rotational velocities.

@@

Figure \ref{fig:muhisto} shows the results for a selection of models
with $\tau_{\rm H}=10^{5}$, different rotational velocities and the two
types of source distributions.
This Figure shows that $F(\mu)$ is consistent with being flat, apart
from some statistical fluctuations on the order of \%2.

@@

Figure \ref{fig:widthvsmu} shows the FWHM for all models as a function
of the viewing angle. 
The FWHM increases for decreasing values of $\mu$ (movement from the
poles to the equator) and increasing values of $V_{\rm max}$. 
In Figure
\ref{fig:widthsvsvelocity} we fix $|\mu|<0.1$, i.e. viewing angle
perpendicular to the rotation axis, to plot the FWHM as a function of
rotational velocity.

@@

All the models fall into a single family of lines in the plane shown
in the right panel of Figure \ref{fig:widthsvsvelocity}, justifying
the choice of our parametrization.
We fit simultaneously all the points in two separate groups, central
and homogeneous sources.
We find that these values are $\lambda_{\rm c} = 0.83 \pm 0.06$ and
$\lambda_{\rm    h}= 0.82\pm 0.05$ respectively.

@@

We measure the peak maxima position, $x_m$, to quantify the transition from
double into single peak profiles. 
In Figure \ref{fig:maximumvsvelocity} we show the dependence of $x_m$ with
the viewing angle parametrized by $|\cos\theta|$ for different
rotational velocities.
There are two interesting features that deserve attention. 
First, for a viewing angle parallel to the rotational axis ($\mu\sim
1.0$) the maxima of all models with the same kind of source
initialization are similar regardless of the rotational velocity.  
Second, at a viewing angle perpendicular to the rotation axis ($\mu\sim 0.0$) a
large fraction of models become single peaked. 
This feature appears more frequently for homogeneously distributed
sources if all the other parameters are equal.

@@

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the cosinus of the
outgoing angle $|\cos\theta|$ and the rotational velocity
$V_{\rm max}$. 
From the right panel observe that the number of scatterings and the
outgoing angle are independent. 
This plot corresponds to the specific case of the central model with
$\tau=10^5$ and $V_{\rm max}=300$\kms, but we have verified that this
holds for all models.

The right panel of Figure \ref{fig:Nscatt} shows how the average
number of scatterings is also independent from the rotational
velocity. 
The lower number of average scatterings in the homogeneous source
distribution is due to a purely geometrical effect.
Photons emitted close to the surface go through less scatterings
before escaping.

%In order to gain a deeper understanding of these results  we make 2D
%histograms for the number of scatterings as a function of the outgoing
%dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show these 
%results in the case $\tau_{\rm H}=10^5$ for the
%static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
%results for the homogeneous (central) source distribution. The color
%scale is logarithmic in the number of photons at a certain value
%$x-N_{\rm scatt}$.

%The top right panel of Figure \ref{fig:Nscatt2D} (homogeneous sources,
%high rotational velocity) supports our hypothesis about
%the photo-sphere in the homogeneous distribution. In this case most of
%the photons that left with $x\sim 0$ have escaped with less than $10$
%scatterings. This also explains the decrease in the average number of
%scatterings observed in Figure \ref{fig:Nscatt}. We have also veryfied
%that these photons are indeed emitted closer to the surface than to
%the center of the sphere.

%However, for a central distribution the situation is quite different
%(lower panels in Figure \ref{fig:Nscatt2D}). In this case the number
%of scatterings remains high, in the order of the optical depth, but
%the two peaks do get closer to each other. The most probable
%physical picture is that each scattering, due to the bulk velocity of
%the gas, is inefficient in driving the photon outside the line
%center.

@@

There are three key results that allows us to build an analytical description to
first order of our results: the independence of the integrated flux,
the number of scatterings and the escape fraction with the rotational
velocity and the viewing angle.

One way to understand these results is that actually a rotating sphere
is not very different from a static one. We can think that in the
frame of the rotating gas cloud all atoms are stationary with respect
to each other, making the scattering process proceed identical as in
the static case.  
While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame.  
In the frame of the rotating gas cloud, all atoms are stationary with
respect to each othe and the scattering process proceeds identical as
in the static case.

An alternative - more quantitative way - to view this is   to consider
the path of an individual photon: say the photon travels  a distance
corresponding to $\tau=1$, scatters by $90$ degrees, then travels
another distance corresponding to $\tau=1$. 
It is easy to show that the point where the photon ends up is
independent of the assumed rotation velocity of the cloud.

@@

Fig~\ref{fig:comparison} shows some examples of analytic vs. full MC
spectra using this approach (the implementation details are in the Appendix).
The left panel shows the results for different rotational velocities
in the case of $\tau_{H}=10^7$ and an observer located perpendicular
to the axis of rotation ($i=0$ in the scheme of Fig~\ref{fig:scheme}
in the Appendix). 
The right panel shows the results for different viewing angles in the
case of $\tau_{H}=10^7$ and a rotational velocity of $V_{\rm
  max}=300$\kms.

@@

{\bf Anisotropic emission}. In our studies the integrated flux is
constant is independent of the viewing angle. This rules out the
possibility of having an observational indication of gas bulk rotation
from measurements of the line equivalent width. A possible way to
obtain an anisotropic flux with the same gas configuration is to allow
for anisotropic emission \citep{Zheng2013,Behrens2014}. We explore
these scenarios in future work (Remolina-Gutierrez et al., in prep.).

@@

\item The single peaked line emerges at viewing angles $\mu\sim 1$ for
  when the rotational velocity is close to than half the FWHM$_0$.

\item  The escape fraction does not change with rotational velocity
  nor with the viewing angle. We interpret this
  in terms of the average number of scatterings which do not 
  significantly change with rotation or viewing angle. This is due to the  
  fact that the vast majority of scatterings events are resonant, 
  during which the mean free path is very short and the effect of gas
  bulk rotation is not enough to affect \ly photons.

\end{itemize}

@@

\begin{displaymath}
J(x,b,\phi,i)=\frac{\sqrt{\pi}}{\sqrt{24}a\tau_0}\Bigg{(}\frac{(x-x_{\rm
    b})^2}{1+{\rm cosh}\Big{[}\sqrt{\frac{2\pi^3}{27}}\frac{|(x
      -x_{\rm b})^3|}{a\tau_0}\Big{]}}\Bigg{)},
\end{displaymath} 
%
where $x_b\equiv v_{b}/v_{\rm th}$, and $v_b$ is the component of ${\bf v}$ projected onto the line-of-sight. This component is given by
\begin{equation}
v_{\rm b}(b,\phi,i)=V_{\rm max}\frac{\sqrt{R^2 -s^2}}{R}\cos i  \hs \cos a,
\end{equation} where the factor $\cos i$ accounts for the projection onto the $x-y$ plane, and the factor $\cos a$ for the subsequent projection onto the line-of-sight. The {\it right panel} of Figure~\ref{fig:scheme} shows that this angle $a$ can be computed from
%
\begin{equation}
\tan [90^{\circ}- a]=\frac{c}{d}=\frac{ b\sin \phi}{\sqrt{R^2 -b^2}}.
\end{equation}

>>>>

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$ 
emission line in distant galaxies. 
We model a galaxy as a sphere with an homogeneous mixture of dust and
hydrogen at a constant temperature. 
These spheres have a solid-body rotation with maximum velocities in
the range $0-300$ \kms and neutral hydrogen optical depths in the
range $\tau_{\rm H}=10^{5}-10^{7}$. 
We consider two types of source distributions in the sphere: central and a
homogeneous.
Our main result is that rotation introduces a dependence of the
line morphology with  viewing angle and rotational velocity. 
Observations with a line of sight parallel to the rotation axis yield
line morphologies similar to the static case. 
For lines of sight perpendicular to the rotation axis both the intensity at the line
center and the line width increase with rotational velocity.   
Along the same line of sight, the line becomes single peaked  at rotational
velocities close to half the line width in the static case. 
Notably, we find that rotation does not induce any significant spatial
anisotropy in the integrated line flux, the escape fraction or the
average number of scatterings. From this, we conclude that the \lya scattering
through a rotating solid-body proceeds identical to the static
case. The only difference is the doppler shift from the different
regions in the sphere that move with respect to the observer. This
allows us to derive an analytic approximation for the viewing-angle
and rotational velocity of the emerging spectrum.
\end{abstract}

@@

For each photon in the simulation we have its initial position inside
the sphere, direction of propagation $\hat{k}_{\rm in}$ and reduced
frequency $x_{\rm in}$. 
The photon's propagation stops once they cross the
surface of the sphere. At this point we store the position, the outgoing direction
of propagation $\hat{k}_{\rm out}$ and the reduced frequency $x_{\rm
  out}$. We now define the angle $\theta$ by $\cos\theta = \hat{k}_{\rm out}\cdot
\hat{k}\equiv \mu$, it is the angle of the outgoing photons with 
respect to the direction of the angular velocity. We use the variable $\mu$ to
study the anisotropy induced by rotation. Fig. \ref{fig:geometry}
shows the geometry of the problem and the important variables.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f3.pdf}
\end{center}
\caption{Same as Fig. \ref{fig:CentralSpec} for \ly photons
  initialized homogeneously throughout the sphere.
    \label{fig:HomSpec}}  
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f4.pdf}
\end{center}
\caption{Shape of the \ly line for different maximum rotational
  velocities for a LoS perpendicular to the rotation axis
  ($|\mu|\sim 0$). The continuous (dashed) line represents the central
  (homogeneous) source distributions.  The panels follow the same
  distribution as in Fig.s \ref{fig:CentralSpec} and \ref{fig:HomSpec}.
    \label{fig:differentvelocities}}  
\end{figure*}

@@

The central result of this paper are summarized in Fig.s
\ref{fig:CentralSpec} and \ref{fig:HomSpec}. 
They show 2D histograms of the escape frequency $x$ and outgoing angle
$\theta$ parameterized by $|\mu|$. 
Taking into account only photons photons around a value
of $|\mu|$ gives us the emission detected by an observer located at an
angle $\theta$ with respect to the rotation axis. 
We have veryfied that the solutions are indeed symmetric with respect
to $\mu=0$.

From these Fig.s we can see that the line properties change with
rotational velocity and depend on the viewing angle $\theta$.  
In the next subsections we quantify the morphology changes with with
velocity, optical depth and viewing angle.  
We characterize the line morphology by its total intensity, the full
width at half maximu, (FWHM) and the location of the peak maxima. 
In order to interpret the
morphological changes in the line we also report the median number of
scatter for each \ly photon in the simulation. 
For the models where dust is included we measure the escape fraction
as a function of rotational velocity.

@@

The first column in both Fig.s \ref{fig:CentralSpec} and
\ref{fig:HomSpec} shows that for the static sphere the line properties
are independent of $|\mu|$, as it is expected due
to the spherical symmetry. 
However, for increasing rotational velocities, at a fixed optical
depth, there are clear signs that this symmetry is broken.

@@

This is clear in  Fig. \ref{fig:differentvelocities} where we
present the different line morphologies for $|\mu|\sim 0$ in the
homogeneous and central configurations.  
The panels have the same distribution as Fig.s \ref{fig:CentralSpec}
and \ref{fig:HomSpec}.    
There are three clear effects on the line morphology as the rotational
velocity increases.  
First, the line broadens; second, the double peaks reduce their intensity; and
third, the intensity at the line centre rises. 
The last two effects are combined to give the impression that the double
peaks are merged into a single one at high rotational velocities.

@@

Fig. \ref{fig:muhisto} shows the results for a selection of models
with $\tau_{\rm H}=10^{5}$, different rotational velocities and the two
types of source distributions.
This Fig. shows that $F(\mu)$ is consistent with being flat, apart
from some statistical fluctuations on the order of \%2.

@@

Fig. \ref{fig:widthvsmu} shows the FWHM for all models as a function
of the viewing angle. 
The FWHM increases for decreasing values of $\mu$ (movement from the
poles to the equator) and increasing values of $V_{\rm max}$. 
In Fig.
\ref{fig:widthsvsvelocity} we fix $|\mu|<0.1$, i.e. viewing angle
perpendicular to the rotation axis, to plot the FWHM as a function of
rotational velocity.

@@

All the models fall into a single family of lines in the plane shown
in the right panel of Fig. \ref{fig:widthsvsvelocity}, justifying
the choice of our parametrization.
We fit simultaneously all the points in two separate groups, central
and homogeneous sources.
We find that these values are $\lambda_{\rm c} = 0.83 \pm 0.06$ and
$\lambda_{\rm    h}= 0.82\pm 0.05$ respectively.

@@

We measure the peak maxima position, $x_m$, to quantify the transition from
double into single peak profiles. 
In Fig. \ref{fig:maximumvsvelocity} we show the dependence of $x_m$ with
the viewing angle parametrized by $|\cos\theta|$ for different
rotational velocities.
There are two interesting features that deserve attention. 
First, for a viewing angle parallel to the rotational axis ($\mu\sim
1.0$) the maxima of all models with the same kind of source
initialization are similar regardless of the rotational velocity.  
Second, at a viewing angle perpendicular to the rotation axis ($\mu\sim 0.0$) a
large fraction of models become single peaked. 
This feature appears more frequently for homogeneously distributed
sources if all the other parameters are equal.

@@

In Fig. \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the cosinus of the
outgoing angle $|\cos\theta|$ and the rotational velocity
$V_{\rm max}$. 
From the right panel observe that the number of scatterings and the
outgoing angle are independent. 
This plot corresponds to the specific case of the central model with
$\tau=10^5$ and $V_{\rm max}=300$\kms, but we have verified that this
holds for all models.

The right panel of Fig. \ref{fig:Nscatt} shows how the average
number of scatterings is also independent from the rotational
velocity. 
The lower number of average scatterings in the homogeneous source
distribution is due to a purely geometrical effect.
Photons emitted close to the surface go through less scatterings
before escaping.

%In order to gain a deeper understanding of these results  we make 2D
%histograms for the number of scatterings as a function of the outgoing
%dimensionless frequency $x$. In Fig. \ref{fig:Nscatt2D} we show these 
%results in the case $\tau_{\rm H}=10^5$ for the
%static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
%results for the homogeneous (central) source distribution. The color
%scale is logarithmic in the number of photons at a certain value
%$x-N_{\rm scatt}$.

%The top right panel of Fig. \ref{fig:Nscatt2D} (homogeneous sources,
%high rotational velocity) supports our hypothesis about
%the photo-sphere in the homogeneous distribution. In this case most of
%the photons that left with $x\sim 0$ have escaped with less than $10$
%scatterings. This also explains the decrease in the average number of
%scatterings observed in Fig. \ref{fig:Nscatt}. We have also veryfied
%that these photons are indeed emitted closer to the surface than to
%the center of the sphere.

%However, for a central distribution the situation is quite different
%(lower panels in Fig. \ref{fig:Nscatt2D}). In this case the number
%of scatterings remains high, in the order of the optical depth, but
%the two peaks do get closer to each other. The most probable
%physical picture is that each scattering, due to the bulk velocity of
%the gas, is inefficient in driving the photon outside the line
%center.

@@

There is a key result of our simulatios that allows us to build an
analytical description for the outgoing spectra.
It is the independence of the following three quantities with the rotational
velocity and the viewing angle: integrated flux, average number of
scatterings and escape fraction.

One way to understand this is that a rotating sphere
is practically identical from a static one. 
In the frame of the rotating gas cloud all atoms are in fact
stationary with respect to each other, making the scattering process
proceed identical as in the static case.  
While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame.  
Therefore, in the frame of the rotating gas cloud all atoms are
stationary with respect to each othe and the scattering process
proceeds identical as in the static case.

An alternative - more quantitative way - to view this is   to consider
the path of an individual photon: say the photon travels  a distance
corresponding to $\tau=1$, scatters by $90$ degrees, then travels
another distance corresponding to $\tau=1$. 
The point where the photon ends up is independent of the assumed
rotation velocity of the cloud.

@@

Fig~\ref{fig:comparison} shows some examples of analytic versus full MC
spectra using this approach (the implementation details are in the Appendix).
The left panel shows the results for different rotational velocities
in the case of $\tau_{H}=10^7$ and an observer located perpendicular
to the axis of rotation ($i=0$ in the scheme of Fig~\ref{fig:scheme}
in the Appendix). 
The right panel shows the results for different viewing angles in the
case of $\tau_{H}=10^7$ and a rotational velocity of $V_{\rm
  max}=300$\kms.

@@

@@

\item The single peaked line emerges at viewing angles $\mu\sim 1$ for
  when the rotational velocity is close to than half the FWHM$_0$.
\end{itemize}

@@

\begin{displaymath}
J(x,b,\phi,i)=\frac{\sqrt{\pi}}{\sqrt{24}a\tau_0}\Bigg{(}\frac{(x-x_{\rm
    b})^2}{1+{\rm cosh}\Big{[}\sqrt{\frac{2\pi^3}{27}}\frac{|(x
      -x_{\rm b})^3|}{a\tau_0}\Big{]}}\Bigg{)},
\end{displaymath} 
%
where $x_b\equiv v_{b}/v_{\rm th}$, and $v_b$ is the component of ${\bf v}$ projected onto the line-of-sight. This component is given by
\begin{equation}
v_{\rm b}(b,\phi,i)=V_{\rm max}\frac{\sqrt{R^2 -s^2}}{R}\cos i  \hs
\cos a,
\end{equation}
where $\beta = 90^{\circ}-a$. The factor $\cos i$ accounts for the projection onto the $x-y$ plane, and the factor $\cos a$ for the subsequent projection onto the line-of-sight. The {\it right panel} of Fig.~\ref{fig:scheme} shows that this angle $a$ can be computed from
%
\begin{equation}
\tan \beta =\tan[90^{\circ}-a]=\frac{c}{d}=\frac{ b\sin \phi}{\sqrt{R^2 -b^2}},
\end{equation}
%
